Supporting Information:\\ Accurate Binding of Sodium and Calcium to a POPC Bilayer by Effective Inclusion of Electronic Polarization

### 2018-03-13 Manuscript/SI.tex
Updates based on reviewers comments

\newpage
\section{Electronic continuum correction applied to CHARMM36 model}
In order to demonstrate that ECC can be applied also to other than Lipid14 model,
we scaled the charges and Lennard-Jones radii of headgroup, glycerol backbone and
carbonyl atoms of the CHARMM36 model of POPC~\cite{klauda10}. We used the
scaling factors optimized for Lipid14 in the main text, i.e., 0.8 for charges and 0.89 for Lennard-Jones
radius. Changes of the headgroup order parameters as a function of added ions
and number density profiles along membrane normal from the ECC-CHARMM model are shown 
in Figs. \ref{scaledCHARMM} and \ref{scaledCHARMMdensity}, respectively,
together with the results from CHARMM36 simulations \cite{catte16,charmmPOPC450mMCaClfiles}
and experiments \cite{akutsu81,altenbach84}. As shown previously \cite{catte16},
the binding of Ca$^{2+}$ is significantly overestimated in CHARMM36 simulations.
In ECC-CHARMM simulations, the Ca$^{2+}$ binding affinity 
is significantly reduced (Fig. \ref{scaledCHARMMdensity}) and the headgroup order parameter response to the added CaCl$_2$
 is in better agreement with experiments (Fig. \ref{scaledCHARMM}).
In line with the results in main text,
Fig. \ref{scaledCHARMMcontacts} also shows faster equilibration time
for Ca$^{2+}$ binding in ECC-CHARMM model than in the original CHARMM36.

In conclusion, applying ECC to CHARMM36 simulation
improved the ion binding affinity to a lipid bilayer, even thought the
scaling factors optimized for Lipid14 were used. Also the area per molecule in
ECC-CHARMM simulation without ions (61 $\AA^2$) is only slightly
lower than the experimental value (64.3 $\AA^2$). 
The optimal scaling factors for CHARMM36 are expected to be different
than for Lipid14 because the initial partial charges are not the same
in different models. Our results indicate that the performance of ion
binding in CHARMM36, and probably in other lipid models as well, could
be further improved by finding the optimal scaling factors for the partial
charges and Lennard-Jones radii.

\newpage

\begin{figure}[!h]
  \centering
  \includegraphics[width=16.0cm]{../Fig/OP.pdf}
  \caption{\label{scaledCHARMM}
    Changes of the headgroup order parameters as a function of NaCl (left) and CaCl$_2$ (right)
    from simulations with CHARMM36 \cite{klauda10}, simulations with ECC-CHARMM parameters and experiments \cite{akutsu81,altenbach84}. 
    CHARMM36 force field parameters for simulations with and without NaCl were taken from CHARMM-GUI \cite{lee15}.
    These were ran 320 ns and 280 ns, respectively, and the first 100 ns was omitted from the analysis.
    Results for CHARMM36 with CaCl$_2$ were calculated from the last 100 ns from the data available at Ref.~\citenum{charmmPOPC450mMCaClfiles}.
    ECC-CHARMM simulations with NaCl and CaCl$_2$ were ran 260 ns and 600 ns, respectively, and the first 100 ns were omitted from the analysis.
    Sodium, calcium, and chloride ions were modeled as ECC-ions in these simulations with parameters from Refs. \citenum{martinek17, Pluharova2014}.
    All the other simulation parameters in simulations in this figure were equal to the available dataset at Ref.~\citenum{charmmPOPC450mMCaClfiles}.
    The temperature in all simulations was set to 310 K.
  }
\end{figure}

>>>>

\newpage
\section{Electronic continuum correction applied to CHARMM36 model}
In order to demonstrate that the ECC method can be applied also to other than Lipid14 model,
we scaled the charges and Lennard-Jones radii of headgroup, glycerol backbone and
carbonyl atoms of the CHARMM36 model of POPC~\cite{klauda10}. We used the
scaling factors optimized for Lipid14 in the main text, i.e., 0.8 for charges and 0.89 for Lennard-Jones
radii. Changes of the headgroup order parameters as a function of added ions
and number density profiles along membrane normal from the ECC-CHARMM model are shown 
in Figs. \ref{scaledCHARMM} and \ref{scaledCHARMMdensity}, respectively,
together with the results from CHARMM36 simulations \cite{catte16,charmmPOPC450mMCaClfiles}
and experiments \cite{akutsu81,altenbach84}. As shown previously \cite{catte16},
the binding of Ca$^{2+}$ is significantly overestimated in CHARMM36 simulations.
In ECC-CHARMM simulations, the Ca$^{2+}$ binding affinity 
is significantly reduced (Fig. \ref{scaledCHARMMdensity}) and the headgroup order parameter response to the added CaCl$_2$
 is in better agreement with experiments (Fig. \ref{scaledCHARMM}).
In line with the results in main text,
Fig. \ref{scaledCHARMMcontacts} also shows faster equilibration time
for Ca$^{2+}$ binding in ECC-CHARMM model than in the original CHARMM36.

In conclusion, applying ECC to CHARMM36 simulation
improved the ion binding affinity to a lipid bilayer, even thought the
scaling factors optimized for Lipid14 were used. Also the area per molecule in
ECC-CHARMM simulation without ions (61 $\AA^2$) is only slightly
lower than the experimental value (64.3 $\AA^2$) \cite{kucerka11}. 
The optimal scaling factors for CHARMM36 are expected to be different
than for Lipid14 because the initial partial charges are not the same
in different models. Our results indicate that the performance of ion
binding in CHARMM36, and probably in other lipid models as well, could
be further improved by finding the optimal scaling factors for the partial
charges and Lennard-Jones radii.

It should be noted that the calcium binding to POPC bilayer in ECC-CHARMM model
is compared here to the CHARMM36 simulation, which was originally ran for
the previous study \cite{catte16,charmmPOPC450mMCaClfiles}.
After the latter simulation has been performed, spesific interaction parameters
between calcium and oxygen \cite{kim16} have been introduced in CHARMM-gui
to reduce the binding of calcium to lipid bilayers. These parameters
reduce the calcium binding to a POPC bilayer to the same level as sodium (data not shown).
Further studies of ion binding with different parameters are in progress
in NMRlipids IV project (\url{nmrlipids.blogspot.fi}), but are beyond the scope of this work.

%\newpage

\begin{figure}[!h]
  \centering
  \includegraphics[width=16.0cm]{../Fig/OP.pdf}
  \caption{\label{scaledCHARMM}
    Changes of the headgroup order parameters as a function of NaCl (left) and CaCl$_2$ (right)
    from simulations with CHARMM36 \cite{klauda10}, simulations with ECC-CHARMM parameters and experiments \cite{akutsu81,altenbach84}. 
    CHARMM36 force field parameters for simulations with and without NaCl were taken from CHARMM-gui \cite{lee15}.
    These were ran 320 ns and 280 ns, respectively, and the first 100 ns was omitted from the analysis.
    Results for CHARMM36 with CaCl$_2$ were calculated from the last 100 ns from the data available at Ref.~\citenum{charmmPOPC450mMCaClfiles}.
    ECC-CHARMM simulations with NaCl and CaCl$_2$ were ran 260 ns and 600 ns, respectively, and the first 100 ns were omitted from the analysis.
    Sodium, calcium, and chloride ions were modeled as ECC-ions in these simulations with parameters from Refs. \citenum{martinek17, Pluharova2014}.
    All the other simulation parameters in simulations in this figure were equal to the available dataset at Ref.~\citenum{charmmPOPC450mMCaClfiles}.
    The temperature in all simulations was set to 310 K.
  }
\end{figure}
